---
title: "homework1"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question #2
Researchers wish to explore the efficacy of triple-drug combinations of antiretroviral therapy for treatment of HIV-infected patients. Because of limitations on potency and the continuing emergence of drug resistance seen with the use of currently available antiretroviral agents in monotherapy and two-drug regimens, triple combination regimens should represent a more promising approach to maximize antiviral activity, maintain long-term efficacy, and reduce the incidence of drug resistance. Towards this end, investigators performed a randomized study comparing AZT + zalcitabine (ddC) versus AZT + zalcitabine (ddC) + saquinavir. The data, time from administration of treatment (in days) until the CD4 count reached a pre-specified level, is given below for the two groups:
AZT + zalcitabine (ddC):
4+, 6, 11, 12, 32, 35, 38+, 39, 45, 49, 75, 80, 84, 85, 87, 102, 180+
AZT + zalcitabine (ddC) + saquinivir:
2, 3, 4, 12, 22, 48, 51+, 56+, 80, 85, 90, 94+, 160, 171, 180, 180+, 238

```{r Q2a,b}
azt_ddc=c("4+",6,11,12,32,35,"38+",39,45,49,75,80,84,85,87,102,"180+")
azt_ddc_saq=c(2,3,4,12,22,48,51,56,80,85,90,94,160,171,180,180,238)

# kmTable=setNames(data.frame(matrix(NA,nrow=1,ncol=length(kmTableColumnNames))),kmTableColumnNames)
kmTable=data.frame()
getKMTable = function(censoredTimesVector,censorSymbol){
  #get numeric representation of censor vector
  censoredTimesVectorNumeric=as.numeric(sub(censorSymbol,'',censoredTimesVector,fixed=TRUE))
  #count number of actual rows in KM table
  cnt_n=length(censoredTimesVectorNumeric)
  #create first row of KM table
  kmTable=setNames(data.frame(matrix(nrow=1,c(0,0,0,cnt_n,as.character("-"),as.character(paste0(cnt_n,"/",cnt_n)),1)),stringsAsFactors=FALSE),c("orderedEventTimes_tj","eventsAtEventTime_ej",
  "censoredObservationsInInterval_cj","inRiskSetAtTime_nj","kaplanMeirSurvivalCurveAtTime_s_tj-1","c_tj-1","kaplanMeirSurvivalCurveAtTime_s_tj"))
  # orderedIndices=order(censoredTimesVectorNumeric)
  # censoredTimesVectorNumeric=censoredTimesVectorNumeric[orderedIndices]
  # censoredTimesVector=censoredTimesVector[orderedIndices]
  censoredTimesVectorNumeric=sort(censoredTimesVectorNumeric)
  for (i in 1:max(censoredTimesVectorNumeric)){
    if(i %in% censoredTimesVectorNumeric){
      #create empty row to fill in
      kmTableRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(kmTable)))),names(kmTable))
      kmTableRow$orderedEventTimes_tj=i
      #count how many events at time
      kmTableRow$eventsAtEventTime_ej=length(which(censoredTimesVector==i))
      #count how many censured at time
      kmTableRow$censoredObservationsInInterval_cj=length(which(censoredTimesVector==paste0(i,censorSymbol)))
      kmTableRow$inRiskSetAtTime_nj=cnt_n
      #sum events and number censored at time
      loss=kmTableRow$eventsAtEventTime_ej+kmTableRow$censoredObservationsInInterval_cj
      kmTableRow[c("kaplanMeirSurvivalCurveAtTime_s_tj-1")]=kmTable[dim(kmTable)[1],c("kaplanMeirSurvivalCurveAtTime_s_tj")]
      kmTableRow[c("c_tj-1")]=paste0((cnt_n-loss),"/",cnt_n)
      kmTableRow$kaplanMeirSurvivalCurveAtTime_s_tj=round((cnt_n-loss)/length(censoredTimesVectorNumeric),3)
      #update count
      cnt_n=cnt_n-loss
      #add row to kmtable
      kmTable=rbind(kmTable,kmTableRow)
    }
  }
  kmTable
}
azt_ddc_KM=getKMTable(azt_ddc,"+")
azt_ddc_saq_KM=getKMTable(azt_ddc_saq,"+")
# write.table(azt_ddc_saq_KM,file="C:/Users/an052283/OneDrive - Cerner Corporation/MSASA/Stat-845/Homeworks/output/Homework1/azt_ddc_saq_KM.tsv",sep="\t",row.names = FALSE)
# write.table(azt_ddc_KM,file="C:/Users/an052283/OneDrive - Cerner Corporation/MSASA/Stat-845/Homeworks/output/Homework1/azt_ddc_KM.tsv",sep="\t",row.names = FALSE)
View(azt_ddc_KM)
View(azt_ddc_saq_KM)

library(survival)
#numeric times and censor list (0 for not censored 1 for censored)
Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))
azt_ddc_KM_R=survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="log-log")
summary(azt_ddc_KM_R)

azt_ddc_saq_KM_R = survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="log-log")
summary(azt_ddc_saq_KM_R)
plot(azt_ddc_KM_R,xlab="Time",ylab="Survival Probability",main="azt_ddc_KM survival curve")
plot(azt_ddc_saq_KM_R,xlab="Time",ylab="Survival Probability",main="azt_ddc_saq_KM survival curve")
```
For both groups separately, construct a data layout (similar to what was done in lecture slides) containing the unique, ordered event times, the number of events that occurred at those unique event times, the number of censored observations in the relevant interval, the number in the risk set at that time, and the Kaplan-Meier estimate of the survival curve at that time. What is the median survival time in the two groups? Will you be comfortable reporting the mean survival time in the two groups?

```{r Q2 c}
#6MP as test case
leukemia_6MP = c(10,7,"32+",23,22,6,16,"34+","32+","25+","11+","20+","19+",6,"17+","35+",6,13,"9+","6+","10+")
getNATable = function(censoredTimesVector,censorSymbol){
  #get numeric representation of censor vector
  censoredTimesVectorNumeric=as.numeric(sub(censorSymbol,'',censoredTimesVector,fixed=TRUE))
  #count number of actual rows in NA table
  cnt_n=length(censoredTimesVectorNumeric)
  #create first row of NA table
  naTable=setNames(data.frame(matrix(nrow=1,c(0,0,cnt_n,0,0,as.character(paste0(cnt_n,"/",cnt_n)),0,0)),stringsAsFactors=FALSE),c("orderedEventTimes_tj","eventsAtEventTime_ej","inRiskSetAtTime_nj","censoredObservationsInInterval_cj","cumulativeHazardRate_ht","d_Y_ratio","cumulativeHazardEstimatedVariance_vt","nelsonAalenSurvivalCurveAtTime_s_tj"))
  censoredTimesVectorNumeric=sort(censoredTimesVectorNumeric)
  sumCensoredInInterval=0
  for (i in 1:max(censoredTimesVectorNumeric)){
    if(i %in% censoredTimesVectorNumeric){
      #create empty row to fill in
      naTableRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(naTable)))),names(naTable))
      naTableRow$orderedEventTimes_tj=i
      #count how many events at time
      naTableRow$eventsAtEventTime_ej=length(which(censoredTimesVector==i))
      #running total of censured between censured time intervals
      naTableRow$censoredObservationsInInterval_cj=length(which(censoredTimesVector==paste0(i,censorSymbol)))
      sumCensoredInInterval=sumCensoredInInterval+naTableRow$censoredObservationsInInterval_cj
      naTableRow$inRiskSetAtTime_nj=cnt_n
      naTableRow$d_Y_ratio=paste0((naTableRow$eventsAtEventTime_ej),"/",cnt_n)
      naTableRow$cumulativeHazardRate_ht=round(as.numeric(naTable[dim(naTable)[1],c("cumulativeHazardRate_ht")])+naTableRow$eventsAtEventTime_ej/cnt_n,3)
      naTableRow$cumulativeHazardEstimatedVariance_vt=round(as.numeric(naTable[dim(naTable)[1],c("cumulativeHazardEstimatedVariance_vt")])+naTableRow$eventsAtEventTime_ej/(cnt_n)^2,3)
      naTableRow$nelsonAalenSurvivalCurveAtTime_s_tj=round(exp(-naTableRow$cumulativeHazardRate_ht),3)
      #sum events and number censored at time
      loss=naTableRow$eventsAtEventTime_ej+naTableRow$censoredObservationsInInterval_cj
      #update count
      cnt_n=cnt_n-loss
      #add row to na table if at least one uncensored variable
      if (i %in% censoredTimesVector){
        naTableRow$censoredObservationsInInterval_cj=sumCensoredInInterval
        naTable=rbind(naTable,naTableRow)
        sumCensoredInInterval=0
      }
    }
  }
  naTable
}

#adapted from http://sas-and-r.blogspot.com/2010/05/example-739-nelson-aalen-estimate-of.html
getCumulativeHazardNA = function(time, event) {
   na.fit = survfit(coxph(Surv(time,event)~1), type="aalen")
   jumps = c(0, na.fit$time, max(time))
   # need to be careful at the beginning and end
   surv = c(1, na.fit$surv, na.fit$surv[length(na.fit$surv)])

   # apply appropriate transformation
   neglogsurv = -log(surv)   
   
   # create placeholder of correct length
   naest = numeric(length(time))  
   for (i in 2:length(jumps)) {
      naest[which(time>=jumps[i-1] & time<=jumps[i])] = 
         neglogsurv[i-1]   # snag the appropriate value
   }
   return(sort(unique(naest)))
}

#TEST SET VALIDATED BY TABLE 4.2 in text no page 95
# leukemia_6MP_NA=getNATable(leukemia_6MP,"+")
# leukemia_6MP_NA_R=survfit(Surv(as.numeric(sub("+","",leukemia_6MP,fixed=TRUE)),ifelse(grepl("+",leukemia_6MP,fixed=TRUE),0,1))~1,conf.type="log-log",type="fh")
# summary(leukemia_6MP_NA_R)
# plot(leukemia_6MP_NA_R,xlab="Time",ylab="Survival Probability",main="leukemia_6MP_NA_R survival curve")
# leukemia_6MP_NA_CH_R=getCumulativeHazardNA(as.numeric(sub("+","",leukemia_6MP,fixed=TRUE)),ifelse(grepl("+",leukemia_6MP,fixed=TRUE),0,1))

azt_ddc_NA=getNATable(azt_ddc,"+")
azt_ddc_NA_R=survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="log-log",type="fh")
summary(azt_ddc_NA_R)
plot(azt_ddc_NA_R,xlab="Time",ylab="Survival Probability",main="azt_ddc_NA survival curve")
#cumulative hazard, confirmation of results
azt_ddc_NA_CH_R=getCumulativeHazardNA(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))
# write.table(azt_ddc_NA,file="C:/Users/an052283/OneDrive - Cerner Corporation/MSASA/Stat-845/Homeworks/output/Homework1/azt_ddc_NA.tsv",sep="\t",row.names = FALSE)
View(azt_ddc_NA)
azt_ddc_saq_NA=getNATable(azt_ddc_saq,"+")
azt_ddc_saq_NA_R = survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="log-log",type="fh")
summary(azt_ddc_saq_NA_R)
plot(azt_ddc_saq_NA_R,xlab="Time",ylab="Survival Probability",main="azt_ddc_saq_NA survival curve")
#cumulative hazard, confirmation of results
azt_ddc_saq_NA_CH_R=getCumulativeHazardNA(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))
View(azt_ddc_saq_NA)
#write.table(azt_ddc_saq_NA,file="C:/Users/an052283/OneDrive - Cerner Corporation/MSASA/Stat-845/Homeworks/output/Homework1/azt_ddc_saq_NA.tsv",sep="\t",row.names = FALSE)

#compare NA to KM survival curve graphically
#azt_ddc
plot(survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="none",type="fh"),xlab="Time",ylab="Survival Probability",main="azt_ddc NA vs. KM survival curves")
lines(survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="none",type="kaplan-meier"),col="red")
legend(150, 1, legend=c("KM", "NA"),
       col=c("red", "black"), lty=1,cex=0.8)
#azt_ddc_saq
plot(survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="none",type="fh"),xlab="Time",ylab="Survival Probability",main="azt_ddc_saq NA vs. KM survival curves")
lines(survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="none",type="kaplan-meier"),col="blue")
legend(150, 1, legend=c("KM", "NA"),
       col=c("blue", "black"), lty=1,cex=0.8)

#compare NA to KM cumulative hazard curve graphically
plot(survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="none",type="fh"),xlab="Time",ylab="Cumulative Hazard",main="azt_ddc NA vs. KM cumulative hazard curves",fun="cumhaz")
lines(survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1,conf.type="none",type="kaplan-meier"),col="red",fun="cumhaz")
legend(150, 1, legend=c("KM", "NA"),
       col=c("red", "black"), lty=1,cex=0.8)
#azt_ddc_saq
plot(survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="none",type="fh"),xlab="Time",ylab="CUmulative Hazard",main="azt_ddc_saq NA vs. KM cumulative hazard curves",fun="cumhaz")
lines(survfit(Surv(as.numeric(sub("+","",azt_ddc_saq,fixed=TRUE)),ifelse(grepl("+",azt_ddc_saq,fixed=TRUE),0,1))~1,conf.type="none",type="kaplan-meier"),col="blue",fun="cumhaz")
legend(150, 1, legend=c("KM", "NA"),
       col=c("blue", "black"), lty=1,cex=0.8)

```

```{r Q2d}
library(KMsurv)
library(tidyverse)
#ninit = length(observations)

#vector of time points, whose length is 1 greater than nlost and nevent
# azt_ddc_c=ifelse(grepl("+",azt_ddc,fixed=TRUE),1,0)

getLifeTableInput = function(censoredTimesVector,censorSymbol){
  #get numeric representation of censor vector
  censoredTimesVectorNumeric=as.numeric(sub(censorSymbol,'',censoredTimesVector,fixed=TRUE))
  #count number of actual rows in KM table
  cnt_n=length(censoredTimesVectorNumeric)
  #create first row of KM table
  lifeTableInputTable=setNames(data.frame(matrix(nrow=1,c(NA,NA,NA)),stringsAsFactors=FALSE),c("time","nlost","nevent"))
  censoredTimesVectorNumeric=sort(censoredTimesVectorNumeric)
  for (i in 1:max(censoredTimesVectorNumeric)){
    if(i %in% censoredTimesVectorNumeric){
      #create empty row to fill in
      lifeTableInputRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(lifeTableInputTable)))),names(lifeTableInputTable))
      lifeTableInputRow$time=i
      #count how many events at time
      lifeTableInputRow$nevent=length(which(censoredTimesVector==i))
      #count how many censured at time
      lifeTableInputRow$nlost=length(which(censoredTimesVector==paste0(i,censorSymbol)))
      lifeTableInputTable=rbind(lifeTableInputTable,lifeTableInputRow)
    }
  }
  na.omit(lifeTableInputTable)
}
azt_ddc_numeric=as.numeric(gsub("+","",azt_ddc,fixed=TRUE))
cuts_ad=seq(0,max(azt_ddc_numeric),60)
#ensure no loss of upper bound when incrementing
if(max(azt_ddc_numeric)>cuts_ad[length(cuts_ad)]){cuts_ad=cuts_ad=c(cuts_ad,(max(azt_ddc_numeric)))}
azt_ddc_lt_raw=getLifeTableInput(azt_ddc,"+")
lifetab_dat=mutate(azt_ddc_lt_raw,time_cat = cut(time, cuts_ad)) %>% group_by(time_cat) %>% summarize(ilost=sum(nlost),ievent=sum(nevent))
azt_ddc_lt=lifetab(tis = c(0,azt_ddc_lt_raw$time), ninit = length(azt_ddc), nlost = azt_ddc_lt_raw$nlost, nevent = azt_ddc_lt_raw$nevent) %>% drop_na(hazard)
azt_ddc_lt_60=lifetab(tis = cuts_ad, ninit = length(azt_ddc), nlost = lifetab_dat$ilost, nevent = lifetab_dat$ievent) 

azt_ddc_saq_numeric=as.numeric(gsub("+","",azt_ddc_saq,fixed=TRUE))
cuts_ads=seq(0,max(azt_ddc_saq_numeric),by=60)
#ensure no loss of upper bound when incrementing
if(max(azt_ddc_saq_numeric)>cuts_ads[length(cuts_ads)]){cuts_ads=c(cuts_ads,(max(azt_ddc_saq_numeric)))}
azt_ddc_saq_lt_raw=getLifeTableInput(azt_ddc_saq,"+")
lifetab_dat=mutate(azt_ddc_saq_lt_raw,time_cat = cut(time, cuts_ads)) %>% group_by(time_cat) %>% summarize(ilost=sum(nlost),ievent=sum(nevent))
azt_ddc_saq_lt=lifetab(tis = c(0,azt_ddc_saq_lt_raw$time), ninit = length(azt_ddc_saq), nlost = azt_ddc_saq_lt_raw$nlost, nevent = azt_ddc_saq_lt_raw$nevent) %>% drop_na(hazard)
azt_ddc_saq_lt_60=lifetab(tis = cuts_ads, ninit = length(azt_ddc_saq), nlost = lifetab_dat$ilost, nevent = lifetab_dat$ievent) 

#plot azt_ddc
plot(cuts_ad,c(0,azt_ddc_lt_60$hazard),type='l',ylab="Hazard function",xlab="days",xaxt="n",main="azt_ddc Life Table Hazard Function in 60 day windows")
axis(1, at = cuts_ad, las=1)
#plot azt_ddc_saq
plot(cuts_ads,c(0,azt_ddc_saq_lt_60$hazard),type='l',ylab="Hazard function",xlab="days",xaxt="n",main="azt_ddc_saq Life Table Hazard Function in 60 day windows")
axis(1, at = cuts_ads, las=1)

```
```{r Q2 e}
library(kmconfband)
#survival function confidence intervals for azt_ddc
azt_ddc_s=survfit(Surv(as.numeric(sub("+","",azt_ddc,fixed=TRUE)),ifelse(grepl("+",azt_ddc,fixed=TRUE),0,1))~1)
azt_ddc_s_ci=summary(azt_ddc_s)
azt_ddc_s_ci_df=data.frame(azt_ddc_s_ci$time,azt_ddc_s_ci$n.risk,azt_ddc_s_ci$n.event,azt_ddc_s_ci$surv,azt_ddc_s_ci$std.err,azt_ddc_s_ci$lower,azt_ddc_s_ci$upper)
#lower and upper intervals and bounds
azt_ddc_s_ci_cb=setNames(cbind(azt_ddc_s_ci_df,confband(azt_ddc_s)[1:dim(azt_ddc_s_ci_df)[1],]),c("time","n.risk","n.event","survival","std.err","lower 95% CI" ,"upper 95% CI","lower 95% CB" ,"upper 95% CB"))
plot(azt_ddc_s,xlab="Time",ylab="Survival Probability",main="azt_ddc KM survival curves")
```
```{r Q2 f}
rtr_example=c(3,4,"5+",6,"6+","8+",11,14,15,"16+")
decrement=0
#redistribute to right value
getRedistributeToRightTable = function(censoredTimesVector,censorSymbol){
  #get numeric representation of censor vector
  censoredTimesVectorNumeric=as.numeric(sub(censorSymbol,'',censoredTimesVector,fixed=TRUE))
  #sort to ensure when determine step# omit correct last element
  censoredTimesVector=censoredTimesVector[order(censoredTimesVectorNumeric)]
  #count number of actual rows in RTR table
  cnt_n=length(censoredTimesVectorNumeric)
  steps=length(which(grepl(censorSymbol,censoredTimesVector[1:length(censoredTimesVector)-1],fixed=TRUE)))
  #create first row of RTR table
  rtrTable=setNames(data.frame(matrix(nrow=1,c(0,(1/cnt_n),"",1)),stringsAsFactors=FALSE),c("observation","decrement","action","S_t"))
  censoredTimesVectorNumeric=sort(censoredTimesVectorNumeric)
  decrement = 0
  base=1/cnt_n
  i=0
  uniqueCensoredTimesVector=unique(censoredTimesVector)
  for (dataPoint in uniqueCensoredTimesVector){
    i=i+1
    #create empty row to fill in
    rtrTableRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(rtrTable)))),names(rtrTable))
    rtrTableRow$observation=dataPoint
    #count how many events or censures at unique dataPoint
    numerator=length(which(censoredTimesVector==dataPoint))
    #censured at time?
    if(length(which(censoredTimesVector==dataPoint & grep(censorSymbol,dataPoint,fixed=TRUE)))>0){
      #no change in survival function
      rtrTableRow$S_t=rtrTable[dim(rtrTable)[1],c("S_t")]
      #display decrement of 0
      rtrTableRow$decrement=0
      #update denominator 
      denominator=length(censoredTimesVector)-max(which(dataPoint == censoredTimesVector))
      if(i==length(uniqueCensoredTimesVector)){
        rtrTableRow$action=paste0("survival is 1-(1)")
        rtrTableRow$S_t=0  
        rtrTableRow$decrement=1
      }
      else{
        rtrTableRow$action=paste0("next time survival is 1-(",decrement," + ",base," + ",base,"*",numerator,"/",denominator,")")
      }
      if(length(uniqueCensoredTimesVector)>=(i+1) && (!grepl(censorSymbol,uniqueCensoredTimesVector[i+1],fixed=TRUE))){
        #update decrement only if not proceeded by a censored observation
        decrement = decrement + base+(base*numerator/denominator)
      }
      #update base
      base=base+(base*numerator/denominator)
    }
    else{
        #update decrement and show action
        if (as.numeric(rtrTable[dim(rtrTable)[1],c("decrement")])==0)
        {
          rtrTableRow$action=paste0("survival is 1-(",decrement,")")
          decrement = decrement
        }
        else{
          decrement = decrement + base
          rtrTableRow$action=paste0("survival is 1-(",decrement,")")
        }
        rtrTableRow$S_t=1-decrement
        #update decrement display
        rtrTableRow$decrement=decrement
    }
      #add row to rtrtable
      rtrTable=rbind(rtrTable,rtrTableRow)
  }
  #eleminate initial bogus row
  rtrTable$decrement=round(as.numeric(rtrTable$decrement),3)
  rtrTable$S_t=round(as.numeric(rtrTable$S_t),3)
  rtrTable[-1,]
}

azt_ddc_saq_RTR=getRedistributeToRightTable(azt_ddc_saq,"+")
azt_ddc_RTR=getRedistributeToRightTable(azt_ddc,"+")
rtr_example_RTR = getRedistributeToRightTable(rtr_example,"+")
```

```{r Q4.7 left truncated methods and data}
#column1 of book page 137 problem 4.7
entry_c1 = c(58,58,59,60,60,61,61,62,62,62,63,63,64,66,66)
exit_c1=c(60,63,69,62,65,72,69,73,66,65,68,74,71,68,69)
death_c1 = c(1,1,0,1,1,0,0,0,1,1,1,0,1,1,1)
#column1 of book page 137 problem 4.7
entry_c2=c(67,67,67,68,69,69,69,70,70,70,71,72,72,73,73)
exit_c2=c(70,77,69,72,79,72,70,76,71,78,79,76,73,80,74)
death_c2=c(1,1,1,1,0,1,1,0,1,0,0,1,1,0,1)
df2_lec = data.frame(cbind(entry_c1,exit_c1,death_c1))
df2_prob = setNames(data.frame(cbind(c(entry_c1,entry_c2),c(exit_c1,exit_c2),c(death_c1,death_c2))),c("entry","exit","death"))

df2_lec_censored_noLT=c(60,63,"69+",62,65,"72+","69+","73+",66,65,68,"74+",71,68,69)
df2_prob_censored_noLT=c(60,63,"69+",62,65,"72+","69+","73+",66,65,68,"74+",71,68,69,70,77,69,72,"79+",72,70,"76+",71,"78+","79+",76,73,"80+",74)

getKM_LT_Table = function(entryExitDeathVector,deathSymbol){
  #create first row of KM_LT table
  km_ltTable=setNames(data.frame(matrix(nrow=1,c(0,0,0,length(entryExitDeathVector),"",1)),stringsAsFactors=FALSE),c("tj","ej","cj","nj","c_tj-1","s_tj"))
  #sort by exit
  orderedIndices=order(entryExitDeathVector$exit)
  entryExitDeathVector=entryExitDeathVector[orderedIndices,]
  for (time in unique(entryExitDeathVector$exit)){
    total_at_risk=length(which(entryExitDeathVector$entry<=time))
    gone=length(which(entryExitDeathVector$exit<time & entryExitDeathVector$death==deathSymbol))
    events=length(which(entryExitDeathVector$exit==time & entryExitDeathVector$death==deathSymbol))
    truncated_at_risk=total_at_risk-gone-as.numeric(km_ltTable$cj[length(km_ltTable$cj)])
    censored=length(which(entryExitDeathVector$exit==time & entryExitDeathVector$death!=deathSymbol))
    numerator=truncated_at_risk-events
    denominator=truncated_at_risk
    #print(paste0(time,": ",numerator,"/",denominator))
    #create empty row to fill in
    km_ltTableRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(km_ltTable)))),names(km_ltTable))
    km_ltTableRow$tj=time
    #count how many events at time
    km_ltTableRow$ej=events
    #count how many censured at time
    km_ltTableRow$cj=censored
    km_ltTableRow$nj=truncated_at_risk
    #sum events and number censored at time
    km_ltTableRow[c("c_tj-1")]=paste0(numerator,"/",denominator)
    km_ltTableRow$s_tj=round((numerator/denominator)*as.numeric(km_ltTable[dim(km_ltTable)[1],c("s_tj")]),4)
    if (km_ltTableRow$ej>0 | km_ltTableRow$s_tj==1){
      #add row to km_lttable
      km_ltTable=rbind(km_ltTable,km_ltTableRow)
    }
  }
  km_ltTable
}
df2_lec_LT=getKM_LT_Table(df2_lec,1)
View(df2_lec_LT)

getKMTableNoCensorRemoval = function(censoredTimesVector,censorSymbol){
  #get numeric representation of censor vector
  censoredTimesVectorNumeric=as.numeric(sub(censorSymbol,'',censoredTimesVector,fixed=TRUE))
  #count number of actual rows in KM table
  cnt_n=length(censoredTimesVectorNumeric)
  #create first row of KM table
  kmTable=setNames(data.frame(matrix(nrow=1,c(0,0,0,cnt_n,as.character("-"),as.character(paste0(cnt_n,"/",cnt_n)),1)),stringsAsFactors=FALSE),c("orderedEventTimes_tj","eventsAtEventTime_ej",
  "censoredObservationsInInterval_cj","inRiskSetAtTime_nj","kaplanMeirSurvivalCurveAtTime_s_tj-1","c_tj-1","kaplanMeirSurvivalCurveAtTime_s_tj"))
  # orderedIndices=order(censoredTimesVectorNumeric)
  # censoredTimesVectorNumeric=censoredTimesVectorNumeric[orderedIndices]
  # censoredTimesVector=censoredTimesVector[orderedIndices]
  censoredTimesVectorNumeric=sort(censoredTimesVectorNumeric)
  for (i in 1:max(censoredTimesVectorNumeric)){
    if(i %in% censoredTimesVectorNumeric){
      #create empty row to fill in
      kmTableRow=setNames(data.frame(matrix(NA,nrow=1,ncol=length(names(kmTable)))),names(kmTable))
      kmTableRow$orderedEventTimes_tj=i
      #count how many events at time
      kmTableRow$eventsAtEventTime_ej=length(which(censoredTimesVector==i))
      #count how many censured at time
      kmTableRow$censoredObservationsInInterval_cj=length(which(censoredTimesVector==paste0(i,censorSymbol)))
      kmTableRow$inRiskSetAtTime_nj=cnt_n
      #sum events and censored
      loss=kmTableRow$eventsAtEventTime_ej+kmTableRow$censoredObservationsInInterval_cj
      kmTableRow[c("kaplanMeirSurvivalCurveAtTime_s_tj-1")]=kmTable[dim(kmTable)[1],c("kaplanMeirSurvivalCurveAtTime_s_tj")]
      #TOOK LAZY WAY OUT AND JUST ADDED BACK IN THE CENSORED OBS - WILL DO CORRECT WAY LATER I Hope
      numerator=(cnt_n-loss+kmTableRow$censoredObservationsInInterval_cj)
      denominator=cnt_n
      kmTableRow[c("c_tj-1")]=paste0(numerator,"/",denominator)
      kmTableRow$kaplanMeirSurvivalCurveAtTime_s_tj=round(numerator/denominator*as.numeric(kmTable[dim(kmTable)[1],c("kaplanMeirSurvivalCurveAtTime_s_tj")]),2)
      #update count
      cnt_n=cnt_n-loss
      #don't add a row when no events 0 should put this at top but no time :0
      if (kmTableRow$eventsAtEventTime_ej>0 | cnt_n==length(censoredTimesVectorNumeric)){
        #add row to kmtable
        kmTable=rbind(kmTable,kmTableRow)
      }
    }
  }
  kmTable
}
df2_lec_censored_noLT_KM=getKMTableNoCensorRemoval(df2_lec_censored_noLT,"+")
View(df2_lec_censored_noLT_KM)
```
4.7(a) Since the diabetics needed to survive long enough from birth until
the study began, the data is left truncated. Construct a table showing
the number of subjects at risk, Y, as a function of age.

```{r 4.7a-c}
#above code and data match output in lecture slide 33 output table  - now try problem data
df2_prob_LT=getKM_LT_Table(df2_prob,1)
View(df2_prob_LT)
```

```{r 4.7d (a-c repeated)}
df2_prob_censored_noLT_KM=getKMTableNoCensorRemoval(df2_prob_censored_noLT,"+")
View(df2_prob_censored_noLT_KM)
```